---
title: "Assignment 4 - Heart rate, respiration and interpersonal coordination"
author: "Riccardo Fusaroli"
date: "August 20, 2019"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 4 - Heart rate, respiration and interpersonal coordination

Physiological data (here heart rate [variability], and respiration) are increasingly popular. Historically treated as pernicious noise to be regressed out of neuro-imaging data, there is now increasing research on how these signals tell us something important about cognition and beyond being just a signal of cognitive processes also impact them in interesting ways. Advanced sport science, and the quantified self movement (closely followed by marketing and communication) have hailed continuous physiological tracking as a powerful way to access and modify attitudes, habits, and performance. Further, as team coordination (in the military, in decision processes and organizational contexts) is more and more in focus, research has attempted to measure how interpersonal coordination between physiological systems might tell us something important about e.g. emotional and cognitive coordination. See references in the reading list for more on this.

In this assignment, you will learn to:
- collect physiological data
- pre-process physiological data (and grow further your mad R skills)
- model the continuous interdependence between two signals (using a multilevel model as proxy for a dynamical system approach)
- conservatively assess the presence of coordination between to signals in a controlled context

This assignment has two parts. The first part familiarizes you with heart rate, and respiration data and their preprocessing. The second part explores how to analyze interpersonal coordination of these signals.

These are the questions you need to be able to answer at the end of the assignment (aka that you need to submit as part of the portfolio)

1) How do you preprocess heart rate and respiration data? Describe the process. If any data needs to be excluded, list the excluded data and motivate the exclusion.

2) Do you observe interpersonal coordination in heart rate and respiration? Describe your control baseline, the method used to quantify coordination, and the statistical models used to infer whether coordination was higher than in the baseline. Report the results of the models.

3) Do you observe differences in coordination between conditions? Report the models and results.

4) Is respiration coordination a likely driver of heart rate coordination? Describe how you would test for it. Bonus points if you actually run the tests and report methods and results.

N.B. to give you a bit more data I included data from previous years (Study1, Study2 and Study 3). Note that synchronouns and turn-taking are the same across both studies, but the third condition is different: in the first year it was self-paced joint reading; in the second year it was the tv-series conversation.

## Let's get started

### Exploring physiological signals

- Choose one pair (one pair, three conditions)
- Load the logs
- Produce a plot of the participants' respiration signal and a different one of the participants' HR signal.
  N.B: remember the slides: artifacts, downsampling, scaling.
  N.B. The gridExtra::grid.arrange() function allows you to display the plots side by side. E.g. grid.arrange(plot1, plot2, plot3, ncol=3). There are also smarter packages, like cowplot and ggpubr.
- Can you eye-ball which condition if any displays more physiological coordination?

### First we read one data file and identify the procedure
```{r}
# Load the libraries
pacman::p_load(tidyverse,gridExtra,groupdata2,reshape2)
```

- Load the file
- correctly identify all columns
- plot the data
- deal with the artifacts
- downsample the dat
- Add a column for study, group, trial and condition

## Try with just one file - in order to construct function
```{r}

# Load the file
one_pair <- read.csv("data/Study1_G1_T1_Synchronous.csv")

# Plot
plot1 <- ggplot(data = one_pair) +
  geom_path(aes(time, Resp1, color = "P1")) +
  geom_path(aes(time, Resp2, color = "P2")) +
  labs(x = "time", y = "Resp") +
  theme(legend.position="bottom")
plot1

## Remove outliers

### Tip, check the function below
removeOuts <- function(ts,threshold){
  ts[ts > (mean(ts,na.rm=T) +
             (threshold*sd(ts,na.rm=T))) | 
       ts < (mean(ts,na.rm=T) -
             (threshold*sd(ts,na.rm=T)))] = mean(ts,na.rm=T)
  return(ts)
}

# threshold=2.5 # Default value at 2.5 sds from the mean

one_pair_extra <- one_pair

# Remove outlieres in respiration
# one_pair_extra$Resp1 <- removeOuts(one_pair_extra$Resp1,2.5)
# one_pair_extra$Resp2 <- removeOuts(one_pair_extra$Resp2,2.5)
# 
# # Remove outlieres in heart rate
# one_pair_extra$HR1 <- removeOuts(one_pair_extra$HR1,2.5)
# one_pair_extra$HR2 <- removeOuts(one_pair_extra$HR2,2.5)

one_pair_extra <-  one_pair_extra %>% mutate_at(vars(Resp1,Resp2,HR1,HR2), removeOuts, threshold = 2.5)


# Run plot on outliere-removed data

# for respiration
plot_resp_outl <- ggplot(data = one_pair_extra) +
  geom_path(aes(time, Resp1, color = "P1")) +
  geom_path(aes(time, Resp2, color = "P2")) +
  labs(x = "Time", y = "Respiration",title = "Respiration (with outlieres removed)") +
  theme(legend.position="bottom")
plot_resp_outl

# for heartrate
plot_HR_outl <- ggplot(data = one_pair_extra) +
  geom_path(aes(time, HR1, color = "P1")) +
  geom_path(aes(time, HR2, color = "P2")) +
  labs(x = "Time", y = "Heart Rate",title = "Heart Rate (with outlieres removed)") +
  theme(legend.position="bottom")
plot_HR_outl

# Plot raw data againt those with the artiacts removed
gridExtra::grid.arrange(plot_HR_outl,plot_resp_outl, plot1)

## Scale
### Tip: if scale() gives some issues, try the one below

# Scale function
z_scale <- function(column){
  column_c <- (column - mean(column)) / sd(column)
}

scaled_data = one_pair_extra %>% mutate_at(vars(Resp1,Resp2,HR1,HR2), z_scale)


# Plot again to check how scaled data look like

# resp scaled
plot_resp_scaled <- ggplot(data = scaled_data) +
  geom_path(aes(time, Resp1, color = "P1")) +
  geom_path(aes(time, Resp2, color = "P2")) +
  labs(x = "Rime", y = "Respiration",title = "Respiration Scaled") +
  theme(legend.position="bottom")
plot_resp_scaled

# HR scaled
plot_HR_scaled <- ggplot(data = scaled_data) +
  geom_path(aes(time, HR1, color = "P1")) +
  geom_path(aes(time, HR2, color = "P2")) +
  labs(x = "Time", y = "Heart Rate", title = "Heart Rate Scaled") +
  theme(legend.position="bottom")
plot_resp_scaled

# Plot all plots together
gridExtra::grid.arrange(plot_resp_outl,plot_HR_outl,plot_resp_scaled,plot_HR_scaled,nrow = 2)

## Downsample
### This is tricky, so you can have a look at my code  (relying on Ludvig's groupdata2) if you get stuck

scaled_data$row <- seq.int(nrow(scaled_data))

data_down = scaled_data %>%
  group(n = 100, method = 'greedy') %>%
  dplyr::summarise(
    time = mean(time,na.rm=T),
    HR1 = mean(HR1,na.rm=T),
    HR2 = mean(HR2,na.rm=T),
    Resp1 = mean(Resp1,na.rm=T),
    Resp2 = mean(Resp2,na.rm=T),
    rowname = row[1]) #the index we use to put them back together 

## Plot the downsampled data
plot_downS <- ggplot(data = data_down) +
  geom_path(aes(time, Resp1, color = "P1")) +
  geom_path(aes(time, Resp2, color = "P2")) +
  labs(x = "time", y = "Resp") +
  theme(legend.position="bottom")
plot_downS

## Now add the group, trial, condition to the cleaned up, scaled, downsampled data
## Tip the info is in the file name
match <- str_match('Study1_G1_T1_Synchronous.csv', "Study([\\d]+)_G([\\d]+)_T([\\d]+)_([:alpha:]+)")
x <- c("file", "study", "group", "trial", "condition")
colnames(match) <- x

# Adding the information from the filename to the dataframe
data_down <- cbind(data_down,match)


```


## Now we are ready to go to load and pre-process all files

Go through all the files (with a function passed onto map_df), check which files should be excluded, if any, and save the pre-processed time-series

A couple of tips:
- looping is oh so slow. Making a function and using Map/Map_df is your salvation.
- each study restarts the group numbering, so you should make sure to change that (e.g. 100 * Study + Group)
- you need to make sure all the data are meaningful or something has to be removed. Plotting is your friend. E.g. "Study1_G1_T1_Synchronous" has one bad respiration signal. We could replace it with NAs


```{r}
### Outliere function
removeOuts <- function(ts,threshold){
  ts[ts > (mean(ts,na.rm=T) +
             (threshold*sd(ts,na.rm=T)))] = mean(ts,na.rm=T) + (threshold*sd(ts,na.rm=T))
  ts[ts < (mean(ts,na.rm=T) -
             (threshold*sd(ts,na.rm=T)))] = mean(ts,na.rm=T) - (threshold*sd(ts,na.rm=T))
  return(ts)
}

### Scale function
z_scale <- function(column){
  column_c <- (column - mean(column)) / sd(column)
}

### Preprocessing function (Define a function running the loading, artifact removal, scaling, downsampling, info adding)
data_preprocess <- function(filename, threshold = 3){
  
  # to be filled in 
  file <- read.csv(filename) #file loading
  
  # change column names to be the same
  colnames(file)[which(names(file) == "min")] <- "time"
  
  # As numeric
  file <- file %>% mutate_at(c("Resp1","Resp2","HR1","HR2","time"), as.numeric)
  
  # removing outlieres based on the threshold in the 4 columns in the list
  d1 <- file %>% mutate_at(c("Resp1","Resp2","HR1","HR2"), removeOuts, threshold)
  
  # Scaling the same 4 columns
  d1 <- d1 %>% mutate_at(c("Resp1","Resp2","HR1","HR2"), z_scale)

  # Downsampling with code from slides
  d1$row <- seq.int(nrow(d1))
  
  d1 = d1 %>%
  group(n = 100, method = 'greedy') %>%
  dplyr::summarise(
    time = mean(time,na.rm=T),
    HR1 = mean(HR1,na.rm=T),
    HR2 = mean(HR2,na.rm=T),
    Resp1 = mean(Resp1,na.rm=T),
    Resp2 = mean(Resp2,na.rm=T),
    rowname = row[1]) 
  
  match <- str_match(filename, "Study([\\d]+)_G([\\d]+)_T([\\d]+)_([:alpha:]+)")
  x <- c("file", "study", "group", "trial", "condition")
  colnames(match) <- x
  
  d1 <- cbind(d1,match)

  # Detecting what data is bad because the belt was loose
  detector1 <- ifelse(duplicated(d1$Resp1)==T,"duplicate","not-duplicate")
  detector2 <- ifelse(duplicated(d1$Resp2)==T,"duplicate","not-duplicate")
  
  if(sum(detector1=="duplicate") >= 100 | sum(detector2=="duplicate") >= 100) {
    d1$Resp1 <- NA
    d1$Resp2 <- NA
  } 

  return(d1)
  
}


#  Identify all files to be read
phys_data = list.files(path = "C:/Users/askes/Dropbox/University/Cognitive Science/3. semester/Experimental Methods 3/portfolio-4/data",pattern = ".csv", full.names = T) %>% purrr::map_df(data_preprocess) 

# Making a unique pair ID
phys_data$study <- as.numeric(phys_data$study)
phys_data$group <- as.numeric(phys_data$group)
phys_data$time <- as.numeric(phys_data$time)

phys_data$uPairID <- (100 * phys_data$study + phys_data$group)

#Normalizing the time values
#Assuming that the large values are millisecond
#choose 400 arbitrarily because it is above a reasonable minute count.
phys_data[which(phys_data$time > 400),]$time <- phys_data[which(phys_data$time > 400),]$time / 1000 / 60
 
#time since 0
phys_data <- phys_data %>% group_by(uPairID, trial) %>% mutate(actual_time_min = time - min(time))

# change timename
colnames(phys_data)[2] <- "time_min"

# Save the data
write.csv(phys_data,"C:/Users/askes/Dropbox/University/Cognitive Science/3. semester/Experimental Methods 3/Portfolio-4/resp_data.csv", row.names = FALSE)


```

## Now we need to run some analysis

Let's start with a multilevel model that accounts for 
- stability (how each signal is autocorrelated)
- interpersonal dependence (each signal is dependent from the previous state of the other signal)

The data needs to be further prepared, so we can analyze both participants in the same model.
We need to turn the data into a long format:
- a column indicating own hr and one own respiration
- a column indicating other hr and one other respiration
- a column indicating change in hr from previous round and one in respiration

We can then run an analysis where change is a function of one's previous state (stability, see slides), and the other's previous state (coupling). Make sure to:
- set up the most interesting contrasts: how do these parameters vary by condition? which condition should be baseline?
- set up the right random effects.
- N.B. the model will be slow. Make sure it works on a subset of the data first!

Bonus question: what if we include an additional layer? Is my heart rate just adjusting to yours, or also to how much you are adjusting to mine?
- to start answering this we can add a column indicating the previous change in hr in the other and one in respiration
- we can then build on the previous models by also adding the previous change in the other



```{r}
# Import data from downloaded csv.
phys_data <- read.csv("resp_data.csv")

# Remove all data in study 3 longer than 3 min.
phys_data$HR1 <- ifelse(phys_data$actual_time_min > 3 & phys_data$study == 3, NA, phys_data$HR1)
phys_data$HR2 <- ifelse(phys_data$actual_time_min > 3 & phys_data$study == 3, NA, phys_data$HR2)
phys_data$Resp1 <- ifelse(phys_data$actual_time_min > 3 & phys_data$study == 3, NA, phys_data$Resp1)
phys_data$Resp2 <- ifelse(phys_data$actual_time_min > 3 & phys_data$study == 3, NA, phys_data$Resp2)

### Remove data for specific files in Heart Rate
list <- c("Study2_G5_T1_Synchronous","Study2_G5_T2_TurnTaking","Study2_G5_T3_Conversation","Study2_G6_T1_TurnTaking","Study2_G6_T2_Conversation","Study2_G6_T3_Synchronous","Study2_G8_T1_TurnTaking","Study2_G8_T2_Synchronous","Study2_G8_T3_Conversation","Study2_G9_T1_Synchronous","Study3_G9_T1_Conversation","Study3_G9_T2_Synchronous","Study3_G9_T3_TurnTaking","Study3_G1_T3_Conversation","Study4_G3_T1_MovementGuided","Study4_G3_T2_MovementCoop","Study4_G3_T3_Synchronous","Study4_G5_T1_Synchronous","Study4_G5_T2_TurnTaking","Study4_G5_T3_Conversation","Study4_G6_T3_Conversation","Study4_G6_T1_TurnTaking","Study4_G6_T2_Synchronous","Study4_G7_T1_MovementGuided","Study4_G7_T2_MovementCoop","Study4_G7_T3_Synchronous","Study4_G8_T4_MovementCoop")

# Apply removal list on data
phys_data$HR1 <- ifelse(phys_data$file %in% list , NA, phys_data$HR1)
phys_data$HR2 <- ifelse(phys_data$file %in% list , NA, phys_data$HR2)

# Genearate a column for each: previous HR1, HR2, Resp1, Resp2. And genearate a column for each: change in HR1, HR2, Resp1, Resp2
phys_data <- phys_data %>% group_by(group,study,trial) %>% 
  mutate(
    
    # For each previous
    HR1_lead = lead(HR1,1),
    HR2_lead = lead(HR2,1),
    Resp1_lead = lead(Resp1,1),
    Resp2_lead = lead(Resp2,1),
    
    # For each change
    change_HR1 = (HR1_lead - HR1),
    change_HR2 = (HR2_lead - HR2),
    change_Resp1 = (Resp1_lead - Resp1),
    change_Resp2 = (Resp2_lead - Resp2))

# Make the data long, so we can analyze both participants at the same time 
## N.B. This is a bit tricky and you might have to do it in several steps

####Long format 

d_hr_lead <- 
  gather(phys_data, # data        
         participant, HR_lead, # new vars
         HR1_lead, HR2_lead) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, HR_lead, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


d_resp_lead <- 
  gather(phys_data, # data        
         participant, Resp_lead, # new vars
         Resp1_lead, Resp2_lead) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, Resp_lead, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


d_hr_change_self <- 
  gather(phys_data, # data        
         participant, HR_change_self, # new vars
         change_HR1, change_HR2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, HR_change_self, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


d_resp_change_self <- 
  gather(phys_data, # data        
         participant, Resp_change_self, # new vars
         change_Resp1, change_Resp2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, Resp_change_self, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))

d_hr_self <- 
  gather(phys_data, # data        
         participant, HR_self, # new vars
         HR1, HR2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, HR_self, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))

d_resp_self <- 
  gather(phys_data, # data        
         participant, Resp_self, # new vars
         Resp1, Resp2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, Resp_self, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))

phys_data_other <- phys_data %>% 
  rename(
    HR1 = "HR2",
    HR2 = "HR1",
    Resp1 = "Resp2",
    Resp2 = "Resp1",
    change_HR1 = "change_HR2",
    change_HR2 = "change_HR1",
    change_Resp1 = "change_Resp2",
    change_Resp2 = "change_Resp1"
  )

d_hr_change_other <- 
  gather(phys_data_other, # data        
         participant, HR_change_other, # new vars
         change_HR1, change_HR2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, HR_change_other, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


d_resp_change_other <- 
  gather(phys_data_other, # data        
         participant, Resp_change_other, # new vars
         change_Resp1, change_Resp2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, Resp_change_other, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


d_hr_other <- 
  gather(phys_data_other, # data        
         participant, HR_other, # new vars
         HR1, HR2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, HR_other, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


d_resp_other <- 
  gather(phys_data_other, # data        
         participant, Resp_other, # new vars
         Resp1, Resp2) %>% #old vars
  
  select( # drop irrelevant vars
    actual_time_min, Resp_other, participant, study, group, condition, trial) %>%
  
  mutate( # create unique participant ID
    participant = parse_number(as.character(group)) * 100 + 
      parse_number(participant))


dd <- cbind(d_hr_change_self,d_hr_change_other,d_resp_change_self,d_resp_change_other,d_hr_lead,d_resp_lead,d_hr_self,d_hr_other,d_resp_self,d_resp_other) 

dd <- dd %>%  select(actual_time_min,participant,group,condition,trial,study,HR_self,Resp_self,HR_other,Resp_other,HR_lead,Resp_lead,HR_change_self,HR_change_other,Resp_change_self,Resp_change_other)



# Plot data
phys_data %>% subset(study==4) %>%
  ggplot() + 
  geom_line(aes(actual_time_min,HR1),color="red") + 
  geom_line(aes(actual_time_min,HR2),color="blue") + 
  facet_grid(group ~ trial)

phys_data %>% subset(study==4) %>%
  ggplot() + 
  geom_line(aes(actual_time_min,Resp1),color="red") + 
  geom_line(aes(actual_time_min,Resp2),color="blue") + 
  facet_grid(group ~ trial)

# Save the data
write.csv(dd,"C:/Users/askes/Dropbox/University/Cognitive Science/3. semester/Experimental Methods 3/Portfolio-4/dd_data.csv", row.names = FALSE)


# Set the most interesting contrast e.g. by defining synchronous or conversation as the baseline
# Model change as a function of own and other previous state 
# Bonus points: Add to the previous model also change in the other to see whether my adaptation is influenced by the other's adaptation.


```




## Now we need to create control baselines.

First shuffled controls, then surrogate pairs.

### Creating controls: shuffled controls

Shuffled controls break the temporal dependencies of time-series by shuffling the value within one time-series. This ensures the "coordination" observed is not due to the actual values in the series and not their sequence.
Tip: sample() is your friend, but make sure to shuffle things within participant/condition and not throughout the whole dataset
 
```{r}

# Create a shuffled dataset
# Concatenate it to the original dataset (and remember to have a column telling you which is which)

# Create the same models as in the previous chunk, but adding an interaction by shuffled vs. real

```
 
 
### TRICKY! Creating controls: surrogate pair controls
 - Per each real pair, identify at least one surrogate pair (matching one of the participants, with somebody doing the same task, but in a different pair)

```{r}

# Identify unique pairs within a given study (to keep things manageable) and create list of possible surrogate pairs (e.g. individual 1 from pair 1 and individual 2 from pair 2)

# Starting from the wide format, create "surrogate" dataset with the data from surrogate pairs

# Make it into long format

# Create models as in chunks above, but adding an interaction with the Real vs. Surrogate variable (exclude shuffled ones for simplicity)



```
 

### Effects of respiration coordination on heart rate coordination
 - describe how you would test those.
 - Optional: run the models and report them

 